---
title: "`r params$title`"
subtitle: "`r params$subtitle`"
date: "`r Sys.Date()`"
format:
  pdf:
    documentclass: article
    papersize: a4
    fontsize: 11pt
    geometry:
      - top=25mm
      - bottom=25mm
      - left=20mm
      - right=20mm
    colorlinks: true
    linkcolor: "ForestGreen"
    toc: true
    toc-depth: 2
    number-sections: true
    fig-pos: "H"
params:
  data_file: ""
  radar_file: ""
  language: "fr"
  title: "Diagnostic Forestier Nemeton"
  subtitle: "Rapport de synthese"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# Load report data
report_data <- readRDS(params$data_file)

# Helper for translations
tr <- function(key) {
  report_data$labels[[key]] %||% key
}

# Color formatting helper
score_color <- function(score) {
  if (score >= 60) "ForestGreen"
  else if (score >= 40) "Orange"
  else "Red"
}
```

# `r tr("project_info")`

**`r tr("project_name")`:** `r report_data$project_name`

`r if (!is.null(report_data$project_description) && nchar(report_data$project_description) > 0) paste0("**Description:** ", report_data$project_description)`

`r if (!is.null(report_data$project_owner) && nchar(report_data$project_owner) > 0) paste0("**", tr("owner"), ":** ", report_data$project_owner)`

- **`r tr("parcels")`:** `r report_data$n_parcels`
- **`r tr("created_at")`:** `r report_data$created_at`

\newpage

# `r tr("global_score_label")`

\begin{center}
\textcolor{`r score_color(report_data$global_score)`}{\Huge\textbf{`r sprintf("%.1f", report_data$global_score)` / 100}}
\end{center}

\vspace{1cm}

## `r tr("radar_title")`

```{r radar}
#| fig-width: 7
#| fig-height: 7
#| fig-align: center

if (file.exists(params$radar_file)) {
  knitr::include_graphics(params$radar_file)
} else {
  cat("Radar plot not available")
}
```

\newpage

# `r tr("family_scores")`

```{r family-table}
#| results: asis

# Build table data
table_data <- do.call(rbind, lapply(report_data$family_stats, function(fam) {
  data.frame(
    Code = fam$code,
    Famille = fam$name,
    Score = fam$mean,
    Min = fam$min,
    Max = fam$max,
    stringsAsFactors = FALSE
  )
}))

# Rename columns based on language
if (params$language == "fr") {
  names(table_data) <- c("Code", "Famille", "Score", "Min", "Max")
} else {
  names(table_data) <- c("Code", "Family", "Score", "Min", "Max")
}

knitr::kable(table_data,
             align = c("c", "l", "c", "c", "c"),
             caption = tr("family_scores"),
             booktabs = TRUE)
```

\vspace{1cm}

## Interpretation des scores

| Score | Interpretation |
|-------|----------------|
| 80-100 | `r if (params$language == "fr") "Excellent - Service ecosystemique tres performant" else "Excellent - Very high ecosystem service performance"` |
| 60-79 | `r if (params$language == "fr") "Bon - Service ecosystemique performant" else "Good - High ecosystem service performance"` |
| 40-59 | `r if (params$language == "fr") "Moyen - Service ecosystemique moyen" else "Average - Moderate ecosystem service performance"` |
| 20-39 | `r if (params$language == "fr") "Faible - Service ecosystemique limite" else "Low - Limited ecosystem service performance"` |
| 0-19 | `r if (params$language == "fr") "Tres faible - Service ecosystemique deficient" else "Very low - Deficient ecosystem service performance"` |

`r if (!is.null(report_data$synthesis_comments) && nchar(report_data$synthesis_comments) > 0) "\\newpage"`

`r if (!is.null(report_data$synthesis_comments) && nchar(report_data$synthesis_comments) > 0) paste0("# ", tr("comments_title"))`

`r if (!is.null(report_data$synthesis_comments) && nchar(report_data$synthesis_comments) > 0) report_data$synthesis_comments`

\vfill

---

*`r tr("generated_by")` v`r packageVersion("nemeton")` - `r Sys.Date()`*
