---
title: "`r params$title`"
subtitle: "`r params$subtitle`"
date: "`r Sys.Date()`"
format:
  pdf:
    documentclass: article
    papersize: a4
    fontsize: 11pt
    geometry:
      - top=25mm
      - bottom=25mm
      - left=20mm
      - right=20mm
    colorlinks: true
    linkcolor: "ForestGreen"
    toc: false
    number-sections: true
    fig-pos: "H"
params:
  data_file: ""
  radar_file: ""
  family_maps_dir: ""
  language: "fr"
  title: "Diagnostic Forestier Nemeton"
  subtitle: "Rapport de synthese"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# Load report data
report_data <- readRDS(params$data_file)

# Helper for translations
tr <- function(key) {
  report_data$labels[[key]] %||% key
}

# Color formatting helper
score_color <- function(score) {
  if (is.na(score)) return("gray")
  if (score >= 60) "ForestGreen"
  else if (score >= 40) "Orange"
  else "Red"
}

# Family order
family_order <- c("C", "B", "W", "A", "F", "L", "T", "R", "S", "P", "E", "N")
```

# `r tr("project_info")`

**`r tr("project_name")`:** `r report_data$project_name`

`r if (!is.null(report_data$project_description) && nchar(report_data$project_description) > 0) paste0("**Description:** ", report_data$project_description)`

`r if (!is.null(report_data$project_owner) && nchar(report_data$project_owner) > 0) paste0("**", tr("owner"), ":** ", report_data$project_owner)`

- **`r tr("parcels")`:** `r report_data$n_parcels`
- **`r tr("created_at")`:** `r report_data$created_at`

\newpage

# `r tr("global_score_label")`

\begin{center}
\textcolor{`r score_color(report_data$global_score)`}{\Huge\textbf{`r sprintf("%.1f", report_data$global_score)` / 100}}
\end{center}

\vspace{1cm}

## `r tr("radar_title")`

```{r radar}
#| fig-width: 7
#| fig-height: 7
#| fig-align: center

if (file.exists(params$radar_file)) {
  knitr::include_graphics(params$radar_file)
} else {
  cat("Radar plot not available")
}
```

\newpage

# `r tr("family_scores")`

```{r family-table}
#| results: asis

# Build table data
table_data <- do.call(rbind, lapply(report_data$family_stats, function(fam) {
  data.frame(
    Code = fam$code,
    Famille = fam$name,
    Score = fam$mean,
    Min = fam$min,
    Max = fam$max,
    stringsAsFactors = FALSE
  )
}))

# Rename columns based on language
if (params$language == "fr") {
  names(table_data) <- c("Code", "Famille", "Score", "Min", "Max")
} else {
  names(table_data) <- c("Code", "Family", "Score", "Min", "Max")
}

knitr::kable(table_data,
             align = c("c", "l", "c", "c", "c"),
             caption = tr("family_scores"),
             booktabs = TRUE)
```

\vspace{1cm}

## `r if (params$language == "fr") "Interpr\u00e9tation des scores" else "Score interpretation"`

| Score | `r if (params$language == "fr") "Interpr\u00e9tation" else "Interpretation"` |
|-------|----------------|
| 80-100 | `r if (params$language == "fr") "Excellent - Service \u00e9cosyst\u00e9mique tr\u00e8s performant" else "Excellent - Very high ecosystem service performance"` |
| 60-79 | `r if (params$language == "fr") "Bon - Service \u00e9cosyst\u00e9mique performant" else "Good - High ecosystem service performance"` |
| 40-59 | `r if (params$language == "fr") "Moyen - Service \u00e9cosyst\u00e9mique moyen" else "Average - Moderate ecosystem service performance"` |
| 20-39 | `r if (params$language == "fr") "Faible - Service \u00e9cosyst\u00e9mique limit\u00e9" else "Low - Limited ecosystem service performance"` |
| 0-19 | `r if (params$language == "fr") "Tr\u00e8s faible - Service \u00e9cosyst\u00e9mique d\u00e9ficient" else "Very low - Deficient ecosystem service performance"` |

`r if (!is.null(report_data$synthesis_comments) && nchar(report_data$synthesis_comments) > 0) "\\newpage"`

`r if (!is.null(report_data$synthesis_comments) && nchar(report_data$synthesis_comments) > 0) paste0("# ", tr("comments_title"))`

```{r}
#| label: synthesis-comments
#| results: asis

# Render markdown comments if available
if (!is.null(report_data$synthesis_comments) && nchar(report_data$synthesis_comments) > 0) {
  cat(report_data$synthesis_comments)
}
```

\newpage

# `r if (params$language == "fr") "Table des mati\u00e8res" else "Table of Contents"`

```{r toc, results='asis'}
# Generate custom TOC
cat("\\tableofcontents\n")
```

\newpage

# `r if (params$language == "fr") "D\u00e9tail par famille d'indicateurs" else "Detail by indicator family"`

```{r family-sections, results='asis'}
for (code in family_order) {
  if (!code %in% names(report_data$family_stats)) next

  fam <- report_data$family_stats[[code]]

  # Section header
  cat("\n\n## ", fam$name, " (", code, ")\n\n", sep = "")

  # Family description
  if (!is.null(report_data$family_descriptions[[code]])) {
    cat(report_data$family_descriptions[[code]], "\n\n")
  }

  # Score summary
  score_label <- if (params$language == "fr") "Score moyen" else "Mean score"
  cat("**", score_label, ":** ", sprintf("%.1f", fam$mean), " / 100", sep = "")
  cat(" (Min: ", sprintf("%.1f", fam$min), ", Max: ", sprintf("%.1f", fam$max), ")\n\n", sep = "")

  # Map image if available
  map_file <- file.path(params$family_maps_dir, paste0("family_", code, "_map.png"))
  if (file.exists(map_file)) {
    cat("### ", if (params$language == "fr") "Carte" else "Map", "\n\n", sep = "")
    # Use relative path from qmd location
    rel_path <- file.path("family_maps", paste0("family_", code, "_map.png"))
    cat("![](", rel_path, "){width=90%}\n\n", sep = "")
  }

  # Indicator details table
  if (!is.null(report_data$indicator_stats[[code]])) {
    cat("### ", if (params$language == "fr") "Indicateurs" else "Indicators", "\n\n", sep = "")

    ind_data <- report_data$indicator_stats[[code]]
    if (nrow(ind_data) > 0) {
      print(knitr::kable(ind_data,
                         align = c("l", "c", "c", "c"),
                         booktabs = TRUE))
      cat("\n\n")
    }
  }

  # Family comments (markdown)
  if (!is.null(report_data$family_comments[[code]]) &&
      nchar(report_data$family_comments[[code]]) > 0) {
    cat("### ", if (params$language == "fr") "Analyse" else "Analysis", "\n\n", sep = "")
    cat(report_data$family_comments[[code]], "\n\n")
  }

  cat("\n\\newpage\n")
}
```

\vfill

---

*`r tr("generated_by")` v`r packageVersion("nemeton")` - `r Sys.Date()`*
